Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2017-03-29 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Jed comments incorporated Abstract through Ch 4

A series of experiments were conducted to characterize a limited set of  residential scale, pre-flashover, or early growth stage, fires.  A fixed fire source footprint, 30.5~cm square, was chosen for this study and a range of common fuels that could be found in and around a residence were selected.  The fires were positioned under an oxygen consumption calorimetry hood to determine the heat release rate.  Heat release rate, flame/plume gas temperatures, and total heat flux from the flames were measured.  In addition, flame movement and height were recorded with photographs and videos.  Given the focus on residential fires, small source fires were selected so that the burn pattern would be limited to less than 2.4~m in height.  Replicate experiments were conducted with each of the fuels, in order to examine the repeatability of the fires and to quantify the variability in the flames.

@@

Three different fuels were used in this study.  Experiments were conducted with a natural gas fueled burner, a liquid fueled pan fire, and a solid fuel. Natural gas was chosen as a fuel for two reasons; it is a common fuel used for heating homes and it is used for calibrating the oxygen depletion calorimeters in NIST’s Large Fire Laboratory.  Therefore the heat content of the gas was monitored and mass flow controlled to provide a heat release rate based on fuel mass flow for comparison with the heat release rate values measured with the calorimeter.  The gas burner arrangement provided the most idealized fire, in that it could be ignited and brought up to a near steady state heat release rate within a few seconds and then be shut down just as quickly.  This provided an approximate `square wave' exposure to the gypsum wallboard.  This is important for determining the potential amount of energy transferred to the exposed surface.

The natural gas burner measured 30.5~cm on each side and the top surface of the burner was elevated 7.5~cm above the floor. The shell of the burner was made from steel and was filled with ``pea sized'' gravel with an average diameter of 6~mm.  A steady state heat release rate of approximately 80 kW was used for the experiments.

@@

Flexible polyurethane foam was chosen as the solid fuel. According to the Polyurethane Foam Association, flexible polyurethane foam is the most widely used cushioning material in upholstered furnishing and mattresses. More than 1.2 billion pounds are produced in the U.S. every year~\cite{Foam,Polyurethane_Foam}. In studies by Ahrens and Evarts of the NFPA, upholstered furniture, and mattresses and bedding were the first items ignited in 33\% of the fatal fires in the U.S. based on averaged data from 2005 through 2009~\cite{Ahrens:2011,Evarts:2011}.

The flexible polyurethane foam had a density of approximately 23~kg/m$^3$.  Each block of foam was 30.5~cm on each side and was 7.5~cm thick. The bottom and sides of the foam block were wrapped in aluminum foil to prevent the fuel from flowing or spreading as it burned to ensure that the fuel area remained with the same fixed area as the gasoline and natural gas.

@@

The source fire characterization experiments were conducted in the NIST Large Fire Research Laboratory's 3~m by 3~m hood oxygen consumption calorimeter. This apparatus follows the methodology of ASTM~E 2067-08, {\em Standard Practice for Full-Scale Oxygen Consumption Calorimetry Fire Tests} ~\cite{ASTM_E2067}. The fires were centered under the hood of the calorimeter and positioned in an open steel frame which provided support for the thermocouples and heat flux sensors. A load cell was positioned on the floor as a support for the fuels.  All of the data was recorded at intervals of 1~s on a computer based data acquisition system.

@@

The NIST Large Fire Research Laboratory's 3~m by 3~m oxygen depletion calorimeter had an estimated peak heat release rate capacity of approximately 700~kW.  Details on the operation and uncertainty in measurements associated with NIST’s 6~m by 6~m oxygen consumption calorimeter were documented by Bryant et al.~\cite{Bryant:2004}. Many of the same instruments were shared between the two calorimeters and calibration burns were conducted during each experimental series.  As a result, the expanded uncertainty of this device has been estimated as 11\% on the measured heat release rate.

@@

Gas temperatures were measured with exposed-bead, Chromel-Alumel (type K) thermocouples, with a 1.0~mm nominal diameter.  Starting from the exposed bead, the thermocouple wire was sheathed in a 3.2~mm diameter inconel shield, 0.76~m in length. The thermocouple was ungrounded. The standard uncertainty in temperature of the thermocouple wire itself is 2.2~$^\circ$C at 277~$^\circ$C and increases to 9.5~$^\circ$C at 871~$^\circ$C as determined by the wire manufacturer.  The response time constant is approximately 0.2~seconds~\cite{Omega}.  The variation of the temperature in the environment surrounding the thermocouple was known to be much greater than that of the wire uncertainty~\cite{Blevins:1999,Pitts:2001}. Small diameter thermocouples were used to limit the impact of radiative heating and cooling.  The estimated combined expanded uncertainty for temperature in these experiments was 15\%.

@@

A 35~mm single lens reflex camera was used to record digital still photographs of the fires.  The camera had an 8 mega-pixel resolution.  A mini DV camcorder with 0.7 mega-pixel resolution was used to capture the video.  A firefighting type thermal imager with 0.08 mega-pixel resolution was used to capture infrared images in the 8 to 14 micron range.  Each camera was positioned on a tripod with a distance of 4.5~m between the centerline of burner to camera lens.  The height of each camera was adjusted so that the center of camera lens was positioned 0.5~m above top of the burner.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{../Figures/Schematic_Drawing}\\
  \caption[Diagram of the source fire flame and plume instrumentation apparatus]{Diagram of the source fire flame and plume instrumentation apparatus which was centered under the 3~m by 3~m oxygen consumption calorimeter.  The heat flux gauges were facing toward the center of the burner/fuel area.}
  \label{Drawing_Fire_Flamer}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{../Figures/Source_Fire_Flamer}\\
  \caption[A photograph of the source fire flame and plume instrumentation apparatus]{A photograph of the source fire flamer and plume instrumentation apparatus which was centered under the 3~m by 3~m hood of the oxygen consumption calorimeter.  In the foreground, from left to right, are the video, 35~mm SLR and thermal imaging camera.}
  \label{Fire_Flamer}
\end{figure}

@@

Figures~\ref{HRR}-\ref{HRR2} show the measured heat release rates for the natural gas, gasoline and polyurethane foam fueled fires. In each graph, the replicate heat release time histories are overlaid to give a sense of the experimental repeatability for each fuel.

@@

On the averaged heat release rate curve for each fuel, the shaded area surrounding the average represents the 95\% confidence level (2$\sigma$), based on Type A evaluation of the uncertainty is provided~\cite{Taylor:1994}.  A Type A evaluation of the uncertainty is any statistical analysis of the data from a series of observations.  In this case the evaluation assumed a normal distribution of the data such that doubling the calculated standard deviation (2$\sigma$) would yield a confidence level of approximately 95\%.

The natural gas burner provided the most repeatable results, with an average steady state heat release rate of 80~kW~$\pm$~9~kW with an average total heat release of 23.3~MJ~$\pm$~1.6~MJ over a 300~s period. The uncertainty of the steady state heat release rate measurement is approximately~$\pm$~11\% It should be noted that 11\% is also the (2$\sigma$) uncertainty estimate of the calorimeter. The gas burner and on/off control valve allowed the heat release rate to be similar in form to a ``square-wave''.

The gasoline pan fire was ignited and then burned until the fuel was consumed.  Therefore the heat release rate curves had more of a ``bell shape''  as opposed to the flat steady- state peak heat release rate region of the natural gas burner.  Therefore the  gasoline average peak heat release rate was based on an average over a 30 s interval bounding the peak heat release rate of each of the 18 experiments.  This method provided an average peak heat release rate of 80~kW~$\pm$~24~kW or 30\%.  The total heat release of the gasoline over the 300~s period after ignition was 13.6~MJ~$\pm$~0.6~MJ.

The block of polyurethane foam demonstrated the largest variability of the three fuels in terms of peak heat release rate and time to peak heat release. Similar to the gasoline that began with a fixed quantity of fuel which was ignited and burned to completion, the polyurethane foam also generated ``bell shaped'' heat release rate curve.  Since the duration of the heat release rate peaks were less than those of the gasoline, the peak heat release rate averaging was performed over a 15~s interval bounding the peak heat release rate from each of the ten experiments.  This process yielded an average peak heat release rate of 45~kW~$\pm$~21~kW or 47\%.  The total heat release over the 300~s period after ignition was 4.1~MJ~$\pm$~0.2~MJ.

The Figures~\ref{HRR}-\ref{HRR2} have a shaded area that shows the 95\% confidence level.  The more narrow the shaded area, the better the repeatability of the fire.  This provides a quick means to assess the repeatability of a given source fire.  Using this measure the natural gas burner fires were the most repeatable, followed by the gasoline pan fires and then the polyurethane fires.

@@

Plots of the centerline temperatures averaged over the full set of replicate experiments for each of the fuels are presented in Fig.~\ref{Temp}. The natural gas temperatures are averaged over a ``steady-state'' burning period.  The temperature differences between the vertical thermocouple locations, for locations greater than 20~cm above the fire location, begin to decrease as the distances from the top of the burner increase.  This trend is similar for both the gasoline fires and the polyurethane foam fires as well.  The most notable difference between the natural gas and the other fuels is the growth and decay of the fire during the test period.

\begin{figure}[p]
  \centering
  \includegraphics[width=.625\columnwidth]{../Figures/FHNG_TC_Plume_Avg}\\
  \includegraphics[width=.625\columnwidth]{../Figures/FHGAS_TC_Plume_Avg}\\
  \includegraphics[width=.625\columnwidth]{../Figures/FHPUF_TC_Plume_Avg}\\
  \caption[Averaged centerline temperature for the natural gas, gasoline, and foam fires]{Averaged centerline temperature for the natural gas, gasoline, and foam fires. The number following the TC represents the height of thermocouple in decimeters above the surface of the burner.}
  \label{Temp}
\end{figure}

@@

Plots of the total heat flux for each of the fuels are presented in Fig.~\ref{Flux}. The repeatability of the total heat flux values varied in a manner similar to the heat release rates, in that the natural gas fires provided a steady state value while the gasoline and the polyurethane foam fires generated a fire curve that has periods of growth and decay.  The average peak total heat flux for the natural gas fires were approximately 2.0~kW/m$^2$.  The average peak values for the gasoline and polyurethane foam were approximately 3.5~kW/m$^2$ and 2.5~kW/m$^2$ respectively.  The values in Fig.~\ref{Flux} are only presented for the experiments which had two heat flux gauges installed at 0.95~m from the centerline of the burners and 0.5~m above the top edge of the burners.

In the initial set of heat release rate experiments another heat flux gauge was used to measure the total heat flux at a location 2~m from the vertical centerline of the burners and 0.5~m above the top of the burners.  At this location the average peak total heat flux measured for the natural gas fires was approximately 0.42~kW/m$^2$.  The average peak values for the gasoline and polyurethane foam were approximately 0.74~kW/m$^2$ and 0.30~kW/m$^2$ respectively. Graphs of the heat flux time histories from the 2~m position are given in Figure~\ref{Flux_2m}

\begin{figure}[p]
  \centering
  \includegraphics[width=.635\columnwidth]{../Figures/HF_NG}\\
  \includegraphics[width=.635\columnwidth]{../Figures/HF_GAS}\\
  \includegraphics[width=.635\columnwidth]{../Figures/HF_PUF}\\
  \caption[Total heat flux versus time for natural gas, gasoline, and foam fires at 0.95~m]{Total heat flux versus time for natural gas, gasoline, and foam fires at a position 0.95~m from the vertical centerline of the burner.}
  \label{Flux}
\end{figure}

\begin{figure}[p]
  \centering
  \includegraphics[width=.635\columnwidth]{../Figures/HF_NG_2m}\\
  \includegraphics[width=.635\columnwidth]{../Figures/HF_Gas_2m}\\
  \includegraphics[width=.635\columnwidth]{../Figures/HF_PUF_2m}\\
  \caption[Total heat flux versus time for natural gas, gasoline, and foam fires at 2~m]{Total heat flux versus time for natural gas, gasoline, and foam fires at a position 2~m from the vertical centerline of the burner.}
  \label{Flux_2m}
\end{figure}

@@

Mass loss measurements were not possible for the natural gas burner experiments, although as noted previously the mass flow of natural gas was approximately 2~g/s.  The mass loss versus time curves for the gasoline and the polyurethane experiments are shown in Fig.~\ref{Mass}.  The gasoline exhibits a linear mass loss rate for the initial 150~s of approximately 2~g/s.  The polyurethane foam was ignited with a small flame in the center of the top surface of the foam block.  As the flame spread across the surface it also began to burn down into the foam sample.  As a result the exposed surface of the foam began to take the shape of a bowl as the burning rate became linear at approximately 2~g/s.

\begin{figure}[p]
  \centering
  \includegraphics[width=4.5in]{../Figures/ML_GAS}\\
  \includegraphics[width=4.5in]{../Figures/ML_PUF}\\
  \caption[Mass loss versus time for the gasoline and foam fueled fires]{Mass loss versus time for the gasoline and foam fueled fires.}
  \label{Mass}
\end{figure}

@@

Each of the experiments was photographed with a digital single lens reflex (SLR) camera, fixed on a tripod, with a capability of capturing at least five images per second.  Located adjacent to the camera was a video camera recording images at approximately thirty 30 frames per second (fps).  The experiments were also recorded with a thermal imager, however the videos were not used for analysis as it was difficult to determine the differences between the visible flame and the thermal plume from the black and white images.

In engineering terms, the ``flame height'' is typically discussed as the median flame height.  A position above the top of the fuel surface where the flame is observed to be present at least 50\% of the time.  As a starting point, the mean flame height for each fuel was estimated from a measurement of the uppermost continuous flame tip from the still photographic images.

Examples of the photographic images are presented in Figs.~\ref{Gas_Foam_Photos} and \ref{Gasoline_Photos}.  Figure~\ref{Gas_Foam_Photos} has photographs of the natural gas flame taken during a period of ``steady state'' burning at approximately 80~kW, the gasoline taken near its average peak heat release rate of 80~kW, and the polyurethane foam fire taken near its peak heat release rate of approximately 60~kW.  Figure~\ref{Gasoline_Photos} presents two photographs of the same gasoline fueled fire when it was burning near its peak heat release rate of 100~kW.  The two photographs, taken 0.2~s apart, represent the range of the observed flame heights from one of the gasoline experiments from 0.65~m and 1.0~m, respectively.

@@

Table~\ref{tab:Flame_Heights} provides the mean visible flame height, as well as the range of flame heights for each fuel, based on analysis of the photographs for three replicate experiments.  In each case, the photographs were chosen for analysis while the fire was generating its peak heat release rate.  The range of flame heights for each fuel is approximately twice the minimum flame height or the continuous flame height. Analyzing the photographs in this manner was very time consuming and was based on making measurements on each photograph with a scale.  Between 6 and 10 seconds of still images, or 30 to 50 frames, were used to approximate the flames heights provided in Table~\ref{tab:Flame_Heights}.  The limit of 5 frames per second also introduces additional uncertainty in determining the flame height. These values will serve as a basis for comparing with mean flame height values determined with video analysis.

As noted by Beyler in his review of fire plume correlations, video tape analysis seems to provide the best estimates of flame height~\cite{Beyler:1986}. Each experiment was recorded with a digital video camera at a refresh rate frequency of 29.97~fps which is the National Television Standards Committee (NTSC) standard. The resulting image in each frame was 720 pixels wide by 480 pixels high. Segments of the video which corresponded to the peak heat release of each experiment were captured with a video editing program.

For the natural gas experiments, approximately 30~s of ``steady state'' video were captured.  In the cases of the gasoline and polyurethane fueled experiments, the video capture period bounded the peak heat release rate.  In the case of the gasoline fueled fires, the peaks occurred over a long period of time, so a 30~s video capture could be made which was representative of the peak burning rate.  The peak duration for the polyurethane foam fires was shorter so, only a 15~s video capture was used.

@@

Each frame from the video clip was saved as a full color Tagged Image File Format (TIFF) file and the pixel aspect ratio was changed from 0.9 to 1, so that each pixel had the same dimension in both the horizontal and vertical directions.  Once these images were rendered by the video editing program, each image was scaled and cropped to provide input for Streams, a suite of image processing programs developed by Nokes for analyzing fluid flow visualization experiments~\cite{Nokes:2011}. The images were scaled so that one pixel equals 5~mm in height and width.

Examples of a set of TIFF images for each fuel are shown in Figures~\ref{flame_height_comp_FHNG80_1} through \ref{flame_height_comp_FHPUF_1}.   Each figure has 56 sequential images from the video segment.  The images total almost 2~s of actual burn time.  Notice how the flame height cycles up and down in each of the sequences.  These images provide additional information and continuity of the motion of the flame so that the analysis is less likely to miss an extreme (minimum or maximum) position when compared to the still photographs.

@@

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/flame_height_comp_FHGAS}\\
	\caption[Example of sequential video frames sampled from a gasoline fueled experiment]{Example of sequential video frames sampled from a gasoline fueled experiment.  In this set of images which spans almost two seconds the visual continuous flame height fluctuates from approximately 0.6~m to 1.1~m.}
	\label{flame_height_comp_FHGAS_1}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/flame_height_comp_FHPUF}\\
	\caption[Example of sequential video frames sampled from a polyurethane foam fueled experiment]{Example of sequential video frames sampled from a polyurethane foam fueled experiment.  In this set of images which spans almost two seconds the continuous visual flame height fluctuates from 0.4~m to 0.9~m.}
	\label{flame_height_comp_FHPUF_1}
\end{figure}

>>>>

A series of experiments were conducted to characterize a limited set of  residential scale, pre-flashover/early growth stage fires.  A fixed fire source footprint, 30.5~cm square, was chosen for this study, and a range of common fuels that could be found in and around a residence were selected.  The fires were positioned under an oxygen consumption calorimetry hood to determine the heat release rate.  Heat release rate, flame/plume gas temperatures, and total heat flux from the flames were measured.  In addition, flame movement and height were recorded with photographs and videos.  Given the focus on residential fires, small source fires were selected so that the burn pattern would be limited to less than 2.4~m in height.  Replicate experiments were conducted with each of the fuels in order to examine the repeatability of the fires and to quantify the variability in the flames.

@@

Three different fuels were used in this study.  Experiments were conducted with a natural gas fueled burner, a liquid fueled pan fire, and a solid fuel. Natural gas was chosen as a fuel for two reasons.  It is a common fuel used for heating homes, and it is used for calibrating the oxygen depletion calorimeters in NIST's Large Fire Laboratory.  Therefore, the heat content of the gas was monitored and mass flow controlled to provide a heat release rate based on fuel mass flow for comparison with the heat release rate values measured with the calorimeter.  The gas burner arrangement provided the most idealized fire, in that it could be ignited and brought up to a near steady state heat release rate within a few seconds, and then be shut down just as quickly.  This provided an approximate square wave exposure to the gypsum wallboard.  This is important for determining the potential amount of energy transferred to the exposed surface.

The natural gas burner measured 30.5~cm on each side, and the top surface of the burner was elevated 7.5~cm above the floor. The shell of the burner was made from steel and was filled with pea-sized gravel with an average diameter of 6~mm.  A steady state heat release rate of approximately 80 kW was used for the experiments.

@@

Flexible polyurethane foam was chosen as the solid fuel. According to the Polyurethane Foam Association, flexible polyurethane foam is the most widely used cushioning material in upholstered furnishings and mattresses. More than 1.2 billion pounds are produced in the U.S. every year~\cite{Foam,Polyurethane_Foam}. In studies by Ahrens and Evarts of the NFPA, upholstered furniture, and mattresses and bedding were the first items ignited in 33\% of the fatal fires in the U.S. based on averaged data from 2005 through 2009~\cite{Ahrens:2011,Evarts:2011}.

The flexible polyurethane foam had a density of approximately 23~kg/m$^3$.  Each block of foam was 30.5~cm on each side and was 7.5~cm thick. The bottom and sides of the foam block were wrapped in aluminum foil to prevent the fuel from flowing or spreading as it burned to ensure that the fuel area remained within the same fixed area as the gasoline and natural gas.

@@

The source fire characterization experiments were conducted in the NIST Large Fire Research Laboratory's 3~m x 3~m hood oxygen consumption calorimeter. This apparatus follows the methodology of ASTM~E 2067-08, {\em Standard Practice for Full-Scale Oxygen Consumption Calorimetry Fire Tests} ~\cite{ASTM_E2067}. The fires were centered under the hood of the calorimeter and positioned in an open steel frame, which provided support for the thermocouples and heat flux sensors. A load cell was positioned on the floor as a support for the fuels.  All of the data was recorded at intervals of 1~s on a computer-based data acquisition system.

@@

The NIST Large Fire Research Laboratory's 3~m x 3~m oxygen depletion calorimeter had an estimated peak heat release rate capacity of approximately 700~kW.  Details on the operation and uncertainty in measurements associated with NIST's 6~m x 6~m oxygen consumption calorimeter were documented by Bryant et al.~\cite{Bryant:2004}. Many of the same instruments were shared between the two calorimeters and calibration burns were conducted during each experimental series.  As a result, the expanded uncertainty of this device has been estimated as 11\% on the measured heat release rate.

@@

Gas temperatures were measured with exposed-bead, Chromel-Alumel (type K) thermocouples, with a 1.0~mm nominal diameter.  Starting from the exposed bead, the thermocouple wire was sheathed in a 3.2~mm diameter inconel shield, 0.76~m in length. The thermocouple was ungrounded. The standard uncertainty in temperature of the thermocouple wire itself is 2.2$^\circ$C at 277$^\circ$C and increases to 9.5$^\circ$C at 871$^\circ$C as determined by the wire manufacturer.  The response time constant is approximately 0.2~seconds~\cite{Omega}.  The variation of the temperature in the environment surrounding the thermocouple was known to be much greater than that of the wire uncertainty~\cite{Blevins:1999,Pitts:2001}. Small diameter thermocouples were used to limit the impact of radiative heating and cooling.  The estimated combined expanded uncertainty for temperature in these experiments was 15\%.

@@

A 35~mm single lens reflex camera was used to record digital still photographs of the fires.  The camera had an 8 mega-pixel resolution.  A mini DV camcorder with 0.7 mega-pixel resolution was used to capture the video.  A firefighting type thermal imager with 0.08 mega-pixel resolution was used to capture infrared images in the 8 to 14 micron range.  Each camera was positioned on a tripod with a distance of 4.5~m between the centerline of burner to camera lens.  The height of each camera was adjusted so that the center of camera lens was positioned 0.5~m above the top of the burner.

@@

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{../Figures/Schematic_Drawing}\\
  \caption[Diagram of the source fire flame and plume instrumentation apparatus]{Diagram of the source fire flame and plume instrumentation apparatus, which was centered under the 3~m x 3~m oxygen consumption calorimeter.  The heat flux gauges were facing toward the center of the burner/fuel area.}
  \label{Drawing_Fire_Flamer}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{../Figures/Source_Fire_Flamer}\\
  \caption[A photograph of the source fire flame and plume instrumentation apparatus]{A photograph of the source fire flamer and plume instrumentation apparatus, which was centered under the 3~m x 3~m hood of the oxygen consumption calorimeter.  In the foreground from left to right are the video, 35~mm SLR, and thermal imaging cameras.}
  \label{Fire_Flamer}
\end{figure}

@@

Figures~\ref{HRR} and \ref{HRR2} show the measured heat release rates for the natural gas, gasoline and polyurethane foam fueled fires. In each graph, the replicate heat release time histories are overlaid to give a sense of the experimental repeatability for each fuel.

@@

On the averaged heat release rate curve for each fuel, the shaded area surrounding the average represents the 95\% confidence level (2$\sigma$), based on Type A evaluation of the uncertainty~\cite{Taylor:1994}.  A Type A evaluation of the uncertainty is any statistical analysis of the data from a series of observations.  In this case, the evaluation assumed a normal distribution of the data such that doubling the calculated standard deviation (2$\sigma$) would yield a confidence level of approximately 95\%.

The natural gas burner provided the most repeatable results, with an average steady state heat release rate of 80~kW~$\pm$~9~kW with an average total heat release of 23.3~MJ~$\pm$~1.6~MJ over a 300~s period. The uncertainty of the steady state heat release rate measurement is approximately~$\pm$~11\% It should be noted that 11\% is also the (2$\sigma$) uncertainty estimate of the calorimeter. The gas burner and on/off control valve allowed the heat release rate to be similar in form to a square-wave.

The gasoline pan fire was ignited and then burned until all the fuel was consumed.  This produced heat release rate curves with more of a bell shape  as opposed to the flat steady- state peak heat release rate region of the natural gas burner.  Therefore, the  gasoline average peak heat release rate was based on an average over a 30 second interval bounding the peak heat release rate of each of the 18 experiments.  This method provided an average peak heat release rate of 80~kW~$\pm$~24~kW, or 30\%.  The total heat release of the gasoline over the 300~seconds period after ignition was 13.6~MJ~$\pm$~0.6~MJ.

The block of polyurethane foam demonstrated the largest variability of the three fuels in terms of peak heat release rate and time to peak heat release. Similar to the gasoline that began with a fixed quantity of fuel which was ignited and burned to completion, the polyurethane foam also generated bell shaped heat release rate curve.  Because the duration of the heat release rate peaks were less than those of the gasoline, the peak heat release rate averaging was performed over a 15~second interval bounding the peak heat release rate from each of the 10 experiments.  This process yielded an average peak heat release rate of 45~kW~$\pm$~21~kW, or 47\%.  The total heat release over the 300~second period after ignition was 4.1~MJ~$\pm$~0.2~MJ.

The Figures~\ref{HRR} and \ref{HRR2} have a shaded area that shows the 95\% confidence level.  The more narrow the shaded area, the better the repeatability of the fire.  This provides a quick means to assess the repeatability of a given source fire.  Using this measure, the natural gas burner fires were the most repeatable, followed by the gasoline pan fires and then the polyurethane fires.

@@

Plots of the centerline temperatures averaged over the full set of replicate experiments for each of the fuels are presented in Fig.~\ref{Temp}. The natural gas temperatures are averaged over a steady state burning period.  The temperature differences between the vertical thermocouple locations, for locations greater than 20~cm above the fire location, begin to decrease as the distances from the top of the burner increase.  This trend is similar for both the gasoline fires and the polyurethane foam fires as well.  The most notable difference between the natural gas and the other fuels is the growth and decay of the fire during the test period.

\begin{figure}[p]
  \centering
  \includegraphics[width=.625\columnwidth]{../Figures/FHNG_TC_Plume_Avg}\\
  \includegraphics[width=.625\columnwidth]{../Figures/FHGAS_TC_Plume_Avg}\\
  \includegraphics[width=.625\columnwidth]{../Figures/FHPUF_TC_Plume_Avg}\\
  \caption[Averaged centerline temperatures for the natural gas, gasoline, and foam fires]{Averaged centerline temperatures for the natural gas, gasoline, and foam fires. The number following the TC represents the height of thermocouple in decimeters above the surface of the burner.}
  \label{Temp}
\end{figure}

@@

Plots of the total heat flux for each of the fuels are presented in Fig.~\ref{Flux}. The repeatability of the total heat flux values varied in a manner similar to the heat release rates, in that the natural gas fires provided a steady state value while the gasoline and the polyurethane foam fires generated a fire curve that features periods of growth and decay.  The average peak total heat flux for the natural gas fires were approximately 2.0~kW/m$^2$.  The average peak values for the gasoline and polyurethane foam were approximately 3.5~kW/m$^2$ and 2.5~kW/m$^2$, respectively.  The values in Fig.~\ref{Flux} are only presented for the experiments that had two heat flux gauges installed at 0.95~m from the centerline of the burners and 0.5~m above the top edge of the burners.

In the initial set of heat release rate experiments, another heat flux gauge was used to measure the total heat flux at a location 2~m from the vertical centerline of the burners and 0.5~m above the top of the burners.  At this location, the average peak total heat flux measured for the natural gas fires was approximately 0.42~kW/m$^2$.  The average peak values for the gasoline and polyurethane foam were approximately 0.74~kW/m$^2$ and 0.30~kW/m$^2$, respectively. Graphs of the heat flux time histories from the 2~m position are given in Figure~\ref{Flux_2m}.

\begin{figure}[p]
  \centering
  \includegraphics[width=.635\columnwidth]{../Figures/HF_NG}\\
  \includegraphics[width=.635\columnwidth]{../Figures/HF_GAS}\\
  \includegraphics[width=.635\columnwidth]{../Figures/HF_PUF}\\
  \caption[Total heat flux vs. time for natural gas, gasoline, and foam fires at 0.95~m]{Total heat flux vs. time for natural gas, gasoline, and foam fires at a position 0.95~m from the vertical centerline of the burner.}
  \label{Flux}
\end{figure}

\begin{figure}[p]
  \centering
  \includegraphics[width=.635\columnwidth]{../Figures/HF_NG_2m}\\
  \includegraphics[width=.635\columnwidth]{../Figures/HF_Gas_2m}\\
  \includegraphics[width=.635\columnwidth]{../Figures/HF_PUF_2m}\\
  \caption[Total heat flux vs. time for natural gas, gasoline, and foam fires at 2~m]{Total heat flux vs. time for natural gas, gasoline, and foam fires at a position 2~m from the vertical centerline of the burner.}
  \label{Flux_2m}
\end{figure}

@@

Mass loss measurements were not possible for the natural gas burner experiments, although as noted previously the mass flow of natural gas was approximately 2~g/s.  The mass loss versus time curves for the gasoline and the polyurethane experiments are shown in Fig.~\ref{Mass}.  The gasoline exhibits a linear mass loss rate for the initial 150~s of approximately 2~g/s.  The polyurethane foam was ignited with a small flame in the center of the top surface of the foam block.  As the flame spread across the surface, it also began to burn down into the foam sample.  As a result the exposed surface of the foam began to take the shape of a bowl as the burning rate became linear at approximately 2~g/s.

\begin{figure}[p]
  \centering
  \includegraphics[width=4.5in]{../Figures/ML_GAS}\\
  \includegraphics[width=4.5in]{../Figures/ML_PUF}\\
  \caption[Mass loss vs. time for the gasoline and foam fueled fires]{Mass loss vs. time for the gasoline and foam fueled fires.}
  \label{Mass}
\end{figure}

@@

Each of the experiments was photographed with a digital single lens reflex (SLR) camera fixed on a tripod, with a capability of capturing at least five still images per second.  Located adjacent to the camera was a video camera recording images at approximately 30 frames per second (fps).  The experiments were also recorded with a thermal imager, but those videos were not used for analysis because it was difficult to determine the differences between the visible flame and the thermal plume from the black and white images.

In engineering terms, flame height is typically discussed as the median flame height.  The median flame height is a position above the top of the fuel surface where the flame is observed to be present at least 50\% of the time.  As a starting point, the mean flame height for each fuel was estimated from a measurement of the uppermost continuous flame tip from the still images.

Examples of the still images are presented in Figs.~\ref{Gas_Foam_Photos} and \ref{Gasoline_Photos}.  Figure~\ref{Gas_Foam_Photos} has photographs of the natural gas flame taken during a period of steady state burning at approximately 80~kW, the gasoline taken near its average peak heat release rate of 80~kW, and the polyurethane foam fire taken near its peak heat release rate of approximately 60~kW.  Figure~\ref{Gasoline_Photos} presents two photographs of the same gasoline fueled fire when it was burning near its peak heat release rate of 100~kW.  The two photographs, taken 0.2~s apart, represent the range of the observed flame heights from one of the gasoline experiments from 0.65~m and 1.0~m, respectively.

@@

Table~\ref{tab:Flame_Heights} provides the mean visible flame height, as well as the range of flame heights for each fuel, based on analysis of the photographs for three replicate experiments.  In each case, the photographs were chosen for analysis while the fire was generating its peak heat release rate.  The range of flame heights for each fuel is approximately twice the minimum flame height or the continuous flame height. Analyzing the photographs in this manner was very time consuming and was based on making measurements on each photograph with a scale.  Between 6 to 10 seconds of still images, or 30 to 50 frames, were used to approximate the flames heights provided in Table~\ref{tab:Flame_Heights}.  The limit of 5 frames per second also introduced additional uncertainty in determining the flame height. These values will serve as a basis for comparing with mean flame height values determined with video analysis.

As noted by Beyler in his review of fire plume correlations, video tape analysis seems to provide the best estimates of flame height~\cite{Beyler:1986}. Each experiment was recorded with a digital video camera at a refresh rate frequency of 29.97~fps, which is the National Television Standards Committee (NTSC) standard. The resulting image in each frame was 720 pixels wide x 480 pixels high. Segments of the video which corresponded to the peak heat release of each experiment were captured with a video editing program.

For the natural gas experiments, approximately 30~seconds of steady state video were captured.  In the cases of the gasoline and polyurethane fueled experiments, the video capture period bounded the peak heat release rate.  In the case of the gasoline fueled fires, the peaks occurred over a long period of time, so a 30~second video capture could be made that was representative of the peak burning rate.  The peak duration for the polyurethane foam fires was shorter, so only a 15~second video capture was used.

@@

Each frame from the video clip was saved as a full color Tagged Image File Format (TIFF) file, and the pixel aspect ratio was changed from 0.9 to 1 so that each pixel had the same dimension in both the horizontal and vertical directions.  Once these images were rendered by the video editing program, each image was scaled and cropped to provide input for Streams, a suite of image processing programs developed by Nokes for analyzing fluid flow visualization experiments~\cite{Nokes:2011}. The images were scaled so that one pixel equals 5~mm in height and width.

Examples of a set of TIFF images for each fuel are shown in Figures~\ref{flame_height_comp_FHNG80_1} through \ref{flame_height_comp_FHPUF_1}.   Each figure has 56 sequential images from the video segment.  The images total almost 2~seconds of actual burn time.  Notice how the flame height cycles up and down in each of the sequences.  These images provide additional information and continuity of the motion of the flame so that the analysis is less likely to miss an extreme (minimum or maximum) position when compared to the still photographs.

@@

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/flame_height_comp_FHGAS}\\
	\caption[Example of sequential video frames sampled from a gasoline fueled experiment]{Example of sequential video frames sampled from a gasoline fueled experiment.  In this set of images, which spans almost two seconds, the visual continuous flame height fluctuates from approximately 0.6~m to 1.1~m.}
	\label{flame_height_comp_FHGAS_1}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/flame_height_comp_FHPUF}\\
	\caption[Example of sequential video frames sampled from a polyurethane foam fueled experiment]{Example of sequential video frames sampled from a polyurethane foam fueled experiment.  In this set of images, which spans almost two seconds, the continuous visual flame height fluctuates from 0.4~m to 0.9~m.}
	\label{flame_height_comp_FHPUF_1}
\end{figure}

From each of the peak heat release rate video clips, the edited TIFF files were loaded into Streams~\cite{Nokes:2011}.  For the natural gas and gasoline fueled fires, approximately 900 images were used in each image sequence.  The polyurethane foam experiments had approximately 450 images in each sequence for each experiment.  Once the image sequence was created, a series of filters and amplifiers are applied to the images to eliminate all colors but red, based on a method developed by Goble~\cite{Goble:2007}.  Once all of the images were converted to a pure red ``flame'' that represented the area of visible fire on a black background, the images were processed within Streams to develop an intensity field based on all of the images being overlaid on each other.  The end result is a contour plot of the intermittency of the flame ranging from zero to one.

The calculations for the intensity field are based on a user-defined computation grid.  A grid sensitivity analysis was conducted with 5~mm, 10~mm, 20~mm, 40~mm, and 80~mm square grids.  Examples of grid cell resolution outputs from Streams for the natural gas fueled burner are shown in Fig.~\ref{Intensity}.  The outputs shown are time-averaged intensity fields generated with the resolutions listed above. In each output the x and y axis dimensions are in millimeters.
